---
title: "CEM via SQL"
author: "Seth Russell"
date: "10/11/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages('bigrquery')
library(bigrquery)
project <- "sandbox-nlp" # put your project ID here
```

## Coarsened Exact Matching using SQL 

Explorations of matching methods for causal inference.

Primary goal is to come up and test SQL statements/functions for matching.

There are two packages that make running queries in BigQuery from R easier:

* __bigQueryR__: http://code.markedmondson.me/bigQueryR/
* __bigrquery__: https://github.com/r-dbi/bigrquery ; also see https://www.tidyverse.org/articles/2018/04/bigrquery-1-0-0/

Alternatively you could use the REST API directly via package such as _httr_

As a sample dataset, we're looking at MIMIC III patients that have a sepsis diagnosis. A positive sepsis diagnosis is defined as having one of the following ICD 9 codes:

"0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"


To see frequency of sepsis diagnoses, see the following query:

```{r sepsis_counts_query}
query <- '
  SELECT icd9_code, count(1) 
  FROM `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD`
  where icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802")
  group by icd9_code
  order by count(1) desc'

# now actually execute query...
gbq_cursor <- bq_project_query(project, query)
sepsis_count_result <- bq_table_download(gbq_cursor)
head(sepsis_count_result, n=10)
```

This next block is just some SQL statements used to figure out the cohort.

```{sql eval=FALSE}
SELECT a.row_id, a.subject_id, a.hadm_id
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left join 
  `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d on
    d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
where d.icd9_code = "0031"
-- should be only 1 record matching criteria


-- this gets the one row and the lowest sequence number (most important sepsis diagnosis)
-- however want to get everyone, not just those with sepsis diagnosis
-- e.g hadm_id 127666 should only show up once - as a sepsis positive admission
 SELECT a.subject_id,
    a.hadm_id,
    d.icd9_code,
    d.seq_num
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
  where icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802")
 and a.subject_id = 1024
 and seq_num = (select min(seq_num) from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dmin where dmin.hadm_id = a.hadm_id and dmin.subject_id = a.subject_id and icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
 

-- get people that have any sepsis diagnosis during their stay.
 SELECT 
    a.subject_id,
    a.hadm_id,
    a.admittime,
    a.dischtime,
    a.admission_type,
    a.admission_location,
    a.insurance,
    a.marital_status,
    a.ethnicity,
    d.icd9_code,
    d.seq_num
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
  where exists (select 'x' from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dd
  where dd.subject_id = a.subject_id and dd.hadm_id = a.hadm_id
  and dd.icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and seq_num = (select min(seq_num) from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dmin where dmin.hadm_id = a.hadm_id and dmin.subject_id = a.subject_id and icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and a.subject_id = 5030
UNION ALL
SELECT distinct 
    a.subject_id,
    a.hadm_id,
    a.admittime,
    a.dischtime,
    a.admission_type,
    a.admission_location,
    a.insurance,
    a.marital_status,
    a.ethnicity,
    '',
    null
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
where not exists (select 'x' from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dd
  where dd.subject_id = a.subject_id and dd.hadm_id = a.hadm_id
  and dd.icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
  and a.subject_id = 5030

 
/* 
 [
  {
    "subject_id": "1024",
    "hadm_id": "127666",
    "icd9_code": "78559",
    "seq_num": "3"
  }
]
*/
 
 
```


??? what treatment to consider ???
how about having a "Percutaneous Transluminal Coronary Angioplasty (PTCA)" procdure:
https://www.qualityindicators.ahrq.gov/Downloads/Modules/IQI/V41/TechSpecs/IQI%2006%20and%20IQI%2030%20Percutaneous%20Transluminal%20Coronary%20Angioplasty.pdf

According to AHRQ, these ICD 9 Procedure Codes indicate a PTCA: "0066", "3601", "3602", "3605"

```{r full_sql_cohort}

query <- '
-- retrieve all sepsis positive patients and if they had the PTCA procedure
-- only about 100 patients that had PTCA that had sepsis
-- first select is those with sepsis but not a PTCA procedure
 SELECT 
    a.subject_id,
    a.hadm_id,
    a.admittime,
    a.dischtime,
    a.admission_type,
    a.admission_location,
    a.insurance,
    a.marital_status,
    a.ethnicity,
    "0" as ptca,
    "1" as sepsis,
    d.icd9_code,
    d.seq_num
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
  where exists (select "x" from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dd
  where dd.subject_id = a.subject_id and dd.hadm_id = a.hadm_id
  and dd.icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and seq_num = (select min(seq_num) from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dmin where dmin.hadm_id = a.hadm_id and dmin.subject_id = a.subject_id and icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and not exists (
  select "x" from `sandbox-nlp.MIMIC3_V1_4.PROCEDURES_ICD` p
  where icd9_code in ("0066", "3601", "3602", "3605")
  and p.subject_id = a.subject_id and p.hadm_id = a.hadm_id
)
UNION ALL
-- this select retrieves those with sepsis and a PTCA procedure
 SELECT 
    a.subject_id,
    a.hadm_id,
    a.admittime,
    a.dischtime,
    a.admission_type,
    a.admission_location,
    a.insurance,
    a.marital_status,
    a.ethnicity,
    "1" as ptca,
    "1" as sepsis,
    d.icd9_code,
    d.seq_num
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
  where exists (select "x" from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dd
  where dd.subject_id = a.subject_id and dd.hadm_id = a.hadm_id
  and dd.icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and seq_num = (select min(seq_num) from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dmin where dmin.hadm_id = a.hadm_id and dmin.subject_id = a.subject_id and icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and exists (
  select "x" from `sandbox-nlp.MIMIC3_V1_4.PROCEDURES_ICD` p
  where icd9_code in ("0066", "3601", "3602", "3605")
  and p.subject_id = a.subject_id and p.hadm_id = a.hadm_id
)
UNION ALL
-- retrieve those without a sepsis diagnosis nor PTCA
 SELECT distinct
    a.subject_id,
    a.hadm_id,
    a.admittime,
    a.dischtime,
    a.admission_type,
    a.admission_location,
    a.insurance,
    a.marital_status,
    a.ethnicity,
    "0" as ptca,
    "0" as sepsis,
    "0" as icd9_code,
    null as seq_num
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
  where not exists (select "x" from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dd
  where dd.subject_id = a.subject_id and dd.hadm_id = a.hadm_id
  and dd.icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and not exists (
  select "x" from `sandbox-nlp.MIMIC3_V1_4.PROCEDURES_ICD` p
  where icd9_code in ("0066", "3601", "3602", "3605")
  and p.subject_id = a.subject_id and p.hadm_id = a.hadm_id
)
UNION ALL
-- retrieve those without a sepsis diagnosis but had a PTCA procedure
 SELECT distinct
    a.subject_id,
    a.hadm_id,
    a.admittime,
    a.dischtime,
    a.admission_type,
    a.admission_location,
    a.insurance,
    a.marital_status,
    a.ethnicity,
    "1" as ptca,
    "0" as sepsis,
    "0" as icd9_code,
    null as seq_num
FROM `sandbox-nlp.MIMIC3_V1_4.ADMISSIONS` a left outer join `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` d
  on d.hadm_id = a.hadm_id and
    d.subject_id = a.subject_id
  where not exists (select "x" from `sandbox-nlp.MIMIC3_V1_4.DIAGNOSES_ICD` dd
  where dd.subject_id = a.subject_id and dd.hadm_id = a.hadm_id
  and dd.icd9_code in ("0031", "0202", "0223", "0270", "0271", "0362", "0380", "03810", "03811", "03812", "03819", "0382", "0383", "03840", "03841", "03842", "03843", "03844", "03849", "0388", "0389", "04082", "0545", "09889", "1125", "77181", "78552", "78559", "78559", "99590", "99591", "99592", "99593", "99594", "99802"))
and exists (
  select "x" from `sandbox-nlp.MIMIC3_V1_4.PROCEDURES_ICD` p
  where icd9_code in ("0066", "3601", "3602", "3605")
  and p.subject_id = a.subject_id and p.hadm_id = a.hadm_id
)
order by subject_id, hadm_id'

gbq_cursor <- bq_project_query(project, query)
cohort <- bq_table_download(gbq_cursor)

prevtopt <- getOption("tibble.width")
options(tibble.width = 110)
prevwopt <- getOption("width")
options(width = 110)
print(cohort)
options(tibble.width = prevtopt)
options(width = prevwopt)
```

## References

"Causal Inference without Balance Checking: Coarsened Exact Matching" https://raw.githubusercontent.com/cran/MatchIt/master/R/matchit2_methods.R

"Multivariate Matching Methods That Are Monotonic Imbalance Bounding" http://gking.harvard.edu/files/gking/files/cem_jasa.pdf

https://gking.harvard.edu/files/political_analysis-2011-iacus-pan_mpr013.pdf